---
title: "Swiss BBS EDA"
author: "Brett Melbourne"
date: "2025-11-12"
output:
    github_document
---

```{r message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
theme_set(theme_bw())
```

Read in the bird data. We'll focus for now on the occurrence data from year 1 in the column `y.1`.

```{r}
willowtit <- read.csv("data/wtmatrix.csv") 
```

Here are the first 30 rows of data

```{r}
willowtit |> 
    select(y.1, elev) |> 
    head(n=30)
```

We see that the data for occurrence are binary with zeros coding for absence and ones for presence. Confirming that we have only zeros and ones:

```{r}
unique(willowtit$y.1)
```

Binary response variables are more challenging to visualize than continuous variables. Plotting the raw zeros and ones data against elevation is a little bit useful but not especially informative (we can see that occurrences appear more often at mid elevation but that's about as good as we can perceive). I've jittered the points and used open circles to help with overlap.

```{r}
willowtit |> 
    ggplot() +
    geom_jitter(mapping=aes(x=elev, y=y.1), shape=1, size=2, height=0.01) +
    labs(x="Elevation (m)", y="Occurrence", 
         title="Occurrence of willow tit in Switzerland")
```

Two useful approaches for visualizing binary data are binning and smoothing. Here, we'll bin elevation into 500 m increments and summarize the binary records of occurrence. We set up bins, then calculate the proportion present in each bin, then the midpoint of the bin. The proportion of occurrences is the sum of the occurrences divided by the number of observations; since the proportion is just the mean of a binary variable, we can use the mean function.

```{r}
freq_table <-
    willowtit |> 
    group_by(bin=cut(elev, breaks=seq(0, 3000, by=500))) |> 
    summarize(p=mean(y.1)) |> 
    mutate(mid=seq(250, 3000, by=500)) 
freq_table
```

Plot including the binned proportion and a smoother on the occurrence data. Since the smoother is modeling the mean of the zeros and ones, it is modeling a smooth function of the proportion of occurrences. The smoother is probably the most useful technique.

```{r}
willowtit |> 
    ggplot() +
    geom_jitter(mapping=aes(x=elev, y=y.1), shape=1, size=2, height=0.01) +
    geom_smooth(mapping=aes(x=elev, y=y.1), linewidth=0.5, se=FALSE) +
    geom_line(data=freq_table, 
              mapping=aes(x=mid, y=p), col=2) +
    geom_point(data=freq_table, 
               mapping=aes(x=mid, y=p), shape=1, size=2, col=2) +
    coord_cartesian(ylim=c(-0.01,1.01)) +
    labs(x="Elevation (m)", y="Occurrence (or proportion)", 
         title="Relationship appears humped")
```

The most striking feature in this plot is that the relationship appears to be hump shaped. There is an optimal elevation for willow tits. Above and below that elevation, the probability of occurrence appears lower.

